Basic concepts and tools for the Toki Pona minimalist language: Wordnet synsets; analysis, synthesis and syntax highlighting of texts

### 2017-12-15 article/article.tex
terminada revisao dos apendices

@@

\appendix
\section{My usage of Toki Pona}\label{mytoki}
I use the standard sounds, but often use [z] for s.
I often translate texts to Toki Pona (e.g. biblical excepts)
and create new texts as poems and short stories.
Most of them are in~\cite{tokisona}.
I omit the li particle after subjects sina and mi,
in accordance with the norm,
but sometimes I use them when there are many predicates.
E.g. sina li wawa li pimeja li lukin pona li moku e kasi mute.
In such cases, the first li is sometimes omitted.
Also, sometimes I use li before mi and sina where I find
that there is unwanted ambiguity, e.g.
sina moku pona e jan 
(might be sina li moku pona e jan or sina moku li pona e jan).

Names are by default transliterated,
but I advocate that, as in other languages,
names might be used as they are in the
correspondent mother tongue.
E.g. the name Erdös is used in
English and Portuguese although the standard
alphabet does not contain ö in such languages.
I also tend to legitimate the use of English (or German) words
in Toki Pona texts if it is the case,
as happens often in scientific writing
(kernel is a German word used in English,
webpage is an English word used in Portuguese).

Proposed notations for numbers seem numerous.
I tend to think that one might indicate is two numbers
are multiplying (pi) or are in different scales
(such as in decimal or binary notations).
For example, I take luka two to mean 52.
Or it might not be taken to such level of strictidness,
for 52 is a reasonable notation for a simple language.
mi jo e jan sama nanpa 12.
Or even:
ona li lon e soweli 27.

Remove the 'li'.
I've been avoiding e ni: and using only :.
've been omitting the subject if it is the
same as in the last sentence.
I've been avoiding also the li sometimes,
and starting only a predicate + object phrase,
or a whole phrase altogether.

pali moku is addiction.
noka and ampa; luka, suno, sike and lawa;
I tend to group these things, as pali and pana.
These clusters of meaning hint me that an even simpler
language is possible and maybe more optimized to the
core goals of Toki Pona.
Also, ambiguities introduced by omission of li
and absence of a token to denote preposition,
suggest to me the possibility of a syntax that is uniquely
parsed.

I am tending to think on a language with the same syntax of toki pona:
<sentence> la <sentence = subject + predicate + object + preposition complement>
but always using the syntax separators,
repeating them when the function in performed in the smallest scale:
'(toki pi toki pona) pi jan sona' (language in toki pona and of intellectuals) vs
'toki pi (toki pona pi pi jan sona)' (language in toki pona of intelectuals).
The keyword for preposition complement is the catch.
An 'a' seems good, but it is taken in Toki Pona.
Maybe use it as the preposition particle and use ha or he
instead of Toki Pona a.
In such a setting,
one might enable pi li e a (concept-qualifier, subject-predicate,
predicate-object, concept-predicate)
in any of the terms of the template:
<sentence> la <sentence = subject + predicate + object + preposition complement>
Between sentences one would use:
la pi, la li, la e, la a, and la la as normally.
This is a very fractal conlang proposal.
Maybe have a way to distinct between noun, qualifiers (adjectives/adverbs) and verb terms,
and accept any of them for subject, predicate, object, predicate,
slots. Or assume a part of speech for template slot or for first word in slot.

Because toki pona's vocabulary is so small,
one might make variations of it for languages
in specific.
For example,
suno might be sun in an English variant,
and sol in a Portuguese variant.
In Appendix~\ref{enPo} are an English and Portuguese
proposition of such variants.
A naive version might be obtained through
choosing the first word in the description of
an official vocabulary, as they exist in English and French.

mute, mute ike,
something can be good, numbered,
or numbered in a bad way, too much,
as in 'o awen, sina pali mute, lape lili'.

% \input{pos}

@@

% jan li ken toki li ken pali pona e toki pona.
toki pona li pona e nasin tan ni:
ona li jo e sitelen mute lili.
ona li pona.
pona kepeken weka 'p' li ona, a.

o taso la toki pona li kalama li lukin pona.
sitelen en nasin li open e sitelen suli\cite{tpLang,pije,fb1,fb2,tokisona,Wikipesija}.
li sona. li nasin e toki e sona e lawa e lon.

@@

ilo lon linha~\ref{hacks} li pana e sitelen
e sona tan sitelen,
en nasin.
linha~\ref{basics} li pana e nasin pi toki pona.
e sitelen tawa kama sona.

mi wile pali e sitelen lon toki pona.
sitelen lon nasin, sitelen sona,
sitelen musi.
ante e toki pona la
la ona li nasa.
taso nasa li pona mute.
li pona mute tawa lawa,
tawa kama sona, tawa sitelen e toki.
ante la toki pona o.
toki e toki pona tawa sina.
kama sona e toki pona sina.

o pona tawa jan pi toki pona.
jan Sonja, Birns-Sprage, Kipo, Pije,
Siwejo, Malija, jan kulupu mute.

@@

>>>>

  \item Make TP variations where each TP word is replaced by a word in
    a language with a large speaker population. 
For example,
suno might be sun in an English variant,
and sol in a Portuguese variant.
A naive version might be obtained through
choosing the first word in the description of
an official vocabulary, as they already exist in English and French.

@@

\appendix
\section{My usage of Toki Pona}\label{mytoki}
I use the standard sounds, but often use [z] for s.
I often translate texts to Toki Pona (e.g. biblical excepts)
and create new texts as poems and short stories.
Most of them are in~\cite{tokisona}
and/or published in the TP Facebook groups.
I omit the li particle after subjects sina and mi,
in accordance with the norm,
but sometimes I use them when there are many predicates.
E.g. sina li wawa li pimeja li lukin pona li moku e kasi mute.
In such cases, the first li is sometimes omitted.
Also, sometimes I use li after mi and sina where I find
that there is unwanted ambiguity, e.g.
sina moku pona 
which 
might be 'sina li moku pona' or 'sina moku pona (li jo e sike)'.

Names are by default transliterated,
but I advocate that, as in other languages,
names might be used as they are in the
correspondent mother tongue.
E.g. the name Erdös is used in
English and Portuguese although the standard
alphabet does not contain ö in such languages.
I also tend to legitimate the use of English (or German) words
in TP texts if it is the case,
as happens often in scientific writing
(kernel is a German word used in English,
webpage and software are English words used in Portuguese).

Proposed notations for numbers seem numerous.
I tend to think that one might indicate if two numbers
are being multiplied (pi) or are in different scales
(such as in decimal or binary notations).
For example, luka (pi) two might mean 52.
Most importantly,
52 is a reasonable notation for a simple language.
E.g.
'mi jo e jan sama nanpa 12',
or
'ona li lon e soweli 27'.

Avoiding needless words...
I've been avoiding e ni: and using only :.
I've been omitting the subject (sometimes also the li)
if it is the
same as in the last sentence.
I tend to group these words:
noka and anpa; luka, suno, sike and lawa; pali and pana.
These clusters of meaning hint me that an even simpler
language is possible and maybe more optimized to the
core goals of TP.

Also, ambiguities introduced by omission of li
and absence of a token to denote preposition,
suggest to me the possibility of a syntax that is always uniquely
parsed (or at least less ambiguous).

I have been inclined to conceive a language with the same syntax of TP:
<sentence> la <sentence = subject + predicate + object (each term with
a possible preposition complement)>
but always using particles to bound the sentence sections,
repeating them when the function is performed in the smallest scale:
'toki pi toki pona pi jan sona' meaning
language in toki pona and of intellectuals vs
'toki pi toki pona pi pi jan sona'
meaning language in [toki pona of intelectuals].
About the keyword for initiating a preposition complement 
might be 'a', but it is already taken in TP.
Maybe use 'a' before a preposition and use 'ha' or 'he'
instead of TP a.
In such a setting,
one might enable pi li e a (concept-qualifier, subject-predicate,
predicate-object, concept-preposition)
in any slot of the template:
<sentence> la <sentence = subject + predicate + object (each term with
a possible preposition complement)>
Between sentences one would use:
la pi, la li, la e, la a, and la la.
This is a very fractal conlang proposal.
Maybe also have a way to discern between a noun,
a qualifier (adjectives/adverbs) and a verb,
and accept any of them for the subject, predicate, object, preposition,
slots. Or assume a part of speech as default for a slot or for the
first word in a slot.

@@

% jan li ken toki li ken pali pona e toki pona.
toki pona li pona e nasin tan ni:
ona li jo e nimi nanpa lili.
ona li pona.
pona kepeken weka 'p' li ona, a.

o taso la toki pona li kalama li lukin pona.
sitelen en nasin li open e sitelen
suli\cite{tpLang,kama,akesiWawa,gdoc,tokisona,Wikipesija}.
li sona. li nasin e toki e sona e lawa e lon.

@@

ilo lon Poki~\ref{hacks} li pana e sitelen
e sona tan sitelen,
en nasin.
Poki~\ref{basics} li pana e nasin pi toki pona.
e sitelen tawa kama sona.

mi wile pali e sitelen lon toki pona.
sitelen lon nasin, sitelen sona,
sitelen musi.
ante e toki pona
la ona li nasa.
taso nasa li pona mute.
li pona mute tawa lawa,
tawa kama sona, tawa sitelen e toki.
ante la toki pona o.
toki e toki pona tawa sina.
kama sona e toki pona sina.

o pona tawa jan pi toki pona.
tawa jan Sonja, Birns-Sprage, Kipo, Pije,
Siwejo, Malija, Tepan, jan kulupu mute.

@@

\bibitem{tokisona}
  Fabbri, R. Toki Sona (a blog dedicated to Toki Pona) (2015-7).
  Available at \url{http://tokisona.github.io/}
